---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# splot - Stefan's plotting package

Just another [ggplot2](http://ggplot2.tidyverse.org) and 
[knitr](https://yihui.name/knitr/) extension package.


## Installation

```{r, eval = FALSE}
# Or the the development version from GitHub:
# install.packages("devtools")
devtools::install_github("stefanedwards/splot")
```



## Usage

We can display a limit on the axes range.
```{r usage1}
library(splot)
ggplot(mtcars, aes(x=cyl, y=mpg)) + 
  geom_point() + 
  coord_capped_cart(bottom='both', left='none') +
  theme_light() + theme(panel.border=element_blank(), axis.line = element_line())
```

**NB**: Disable `panel.border` and enable `axis.line` in `theme`, otherwise you will not see an effect!

We could also show that the x-axis is categorical (or ordinal):
```{r brackets_demo}
ggplot(mtcars, aes(x=as.factor(cyl), y=mpg)) + 
  geom_point(position=position_jitter(width=0.1)) + 
  coord_flex_cart(bottom=brackets_horisontal(), left=capped_vertical('none')) +
  theme_light() + theme(panel.border=element_blank(), axis.line = element_line())
```


## Extensions to knitr

We automatically load knitr's `lnit_print` for 
data frames and dplyr tables to provide automatic pretty printing of
these using `kable`:

Before loading `splot` package:
```{r knit_print_disable,include=FALSE}
options(knit_print=FALSE)
```
```{r}
data(USArrests)
head(USArrests)
```

After loading `splot`:
```{r knit_print_enable,include=FALSE}
options(knit_print=TRUE)
```
```{r}
head(USArrests)
```

See `knit_print.data.frame`.


