---
title: "Capping axis lines"
author: "Stefan McKinnon Edwards <sme@@iysik.com>"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Demonstration

A quick demonstration of capping the lines.

First, load the package, generate a dataset and display it.

```{r}
#library(ggplot2)
library(splot)

dat1 <- data.frame(
  gp = factor(rep(letters[1:3], each = 10)),
  y = rnorm(30),
  cl = sample.int(3, 30, replace=TRUE),
  cl2 = sample(c('a','b','c'), 30, replace=TRUE)
)

my.theme <- theme_light()

(
  p <- ggplot(dat1, aes(gp, y)) + geom_point() + my.theme
)
```

**NB:** In order to manipulate the axis lines, they must be drawn.
Modify the theme so the `panel.border` is not drawn (it will be on top of the axis lines),
and have the axis lines drawn:

```{r}
my.theme <- my.theme + theme(panel.border=element_blank(), axis.line = element_line())
p <- p + my.theme
```

Now, let's have some fun:

```{r}
p + coord_capped_cart(bottom='right', left='none')
p + coord_flex_cart(bottom=capped_horisontal('none'), left=brackets_vertical())
p + coord_capped_cart(bottom=brackets_horisontal())
```


Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Facets

```{r eval=FALSE}

# Facet grid -----------------
ggplot(dat1, aes(gp, y)) + geom_point() + 
  coord_capped_flip(bottom = 'left', left='none') + 
  theme(axis.title=element_blank(), plot.title=element_text(size=rel(1))) +
  facet_rep_grid(~cl)

dat2 <- rbind(dat1, data.frame(gp=letters[1:3], y=rnorm(3, 2), cl=3, cl2='b'))
ggplot(dat2, aes(gp, y)) + geom_point() + 
  coord_capped_flip(bottom = 'left', left='none') + 
  theme(axis.title=element_blank(), plot.title=element_text(size=rel(1))) +
  facet_rep_grid(.~cl, scales='free_y')


```

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
