---
title: "Capping axis lines"
author: "Stefan McKinnon Edwards <sme@iysik.com>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Capping axis lines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,include=FALSE}
library(knitr)

knitr::opts_chunk$set(fig.height=4, fig.width=6)
```

## Demonstration

A quick demonstration of capping the lines.

First, load the package, generate a dataset and display it.

```{r load_pkg_and_data,fig.cap='Default ggplot2 plotting.'}
library(ggplot2)
library(lemon)

dat1 <- data.frame(
  gp = factor(rep(letters[1:3], each = 10)),
  y = rnorm(30),
  cl = sample.int(3, 30, replace=TRUE),
  cl2 = sample(c('a','b','c'), 30, replace=TRUE)
)

my.theme <- theme_light()

(
  p <- ggplot(dat1, aes(gp, y)) + geom_point() + my.theme
)
```

**NB:** In order to manipulate the axis lines, they must be drawn.
Modify the theme so the `panel.border` is not drawn (it will be on top of the axis lines),
and have the axis lines drawn:

```{r theme}
my.theme <- my.theme + theme(panel.border=element_blank(), axis.line = element_line())
p <- p + my.theme
```

Now, let's have some fun.

We cap the bottom axis line to the right-most tick. The left end is also capped by the amount
specified with the `gap` argument (at time of writing, defaulted at `0.01`).
```{r fig.cap='Using `coord_capped_cart` to cap the bottom axis from the right. The left axis is unaffected.'}
p + coord_capped_cart(bottom='right')
```

To keep the axis lines consistent, we also specify the `left` argument, which still caps the left
axis line by the amount specified with the `gap` argument.
```{r fig.cap='As before, but left axis is now also capped to give a consistent look.'}
p + coord_capped_cart(bottom='right', left='none')
```


To avoid overplotting, we can apply a jitter. 
To emphasise that the x-axis is categorical, we can place brackets.
We finally polish the plot by removing the redundant vertical grid lines.
```{r fig.cap='Placing brackets brackets instead of ticks emphasises that the x-scale is categorical and not nominal.'}
ggplot(dat1, aes(gp, y)) + geom_point(position=position_jitter(width=0.2, height=0)) +
  coord_capped_cart(left='none', bottom=brackets_horisontal()) +
  my.theme + theme(panel.grid.major.x = element_blank())
```


## The `coord` objects

ggplot2's Cartesian coordinates systems, `coord_cartesian`, have been extended 
to allow for flexible specification of how axes are drawn. 
You've seen them above. 
The following table summarises the connection between ggplot2's coord functions
and those of splot!.

 ggplot2          | splot's flexible   | splot's short hand
------------------|--------------------|--------------------
`coord_cartesian` | `coord_flex_cart`  | `coord_capped_cart`
`coord_flip`      | `coord_flex_flip`  | `coord_capped_flip`
`coord_fixed`     | `coord_flex_fixed` | --

The short hand functions in the table's right column simply are almost identical
to those in the middle column.
If one of the side arguments are specified with a character value, the 
relevant axis drawing function is used. You can however choose to use e.g.
`brackets_horisontal` in place.

### The axis drawing functions

The functions `capped_horisontal` or `brackets_vertical` returns a function
that is called when ggplot2 prints the plot. It is called with the arguments

    scale_details, axis, scale, position, theme

and the function should return a grob.
Some pointers to how it is used can be found in ggplot2's help pages; just run
`?"ggplot2-ggproto"`. 
In this package, we use ggplot2 to build the axes, then modify in place the
return grobs.

### Brackets

The brackets comes in two orientations: 
`brackets_horisontal` and `brackets_vertical`. 
If you attempt to use a vertical bracket on a horisontal axis, it will fail with
a undescriptive error.

The bracket functions accept a `direction` argument, which can be used to control
which direction the end-points are pointing:

```{r brackets}
p <- ggplot(mpg, aes(cyl, hwy, colour=class)) +
  geom_point(position=position_jitter(width=0.3)) +
  scale_x_continuous(breaks=c(4,5,6,8), sec.axis=dup_axis()) + 
  scale_y_continuous(sec.axis=dup_axis()) +
  coord_flex_cart(bottom=brackets_horisontal(), 
                  top=brackets_horisontal(direction='down'),
                  left=brackets_vertical(), 
                  right=brackets_vertical(direction='right')) +
  my.theme
p
```

The look of the brackets are controlled via `theme(axis.ticks)`. 
The length of the end-point are controlled via the theme 
`theme(axis.ticks.length)`. If these needs to be specified for each margin,
use the argument `tick.length`:

```{r}
p <- ggplot(mpg, aes(cyl, hwy, colour=class)) +
  geom_point(position=position_jitter(width=0.3)) +
  coord_flex_cart(bottom=brackets_horisontal(tick.length=unit(0, 'cm')), 
                  left=brackets_vertical(tick.length = unit(1, 'cm'))) +
  my.theme
p
```

As shown above, using `tick.length=unit(0, 'cm')` results in a flat line.

## Facets

Having produced such wonderous axes, it is a pity they are not plotted around 
all panels when using faceting. 

```{r}
dsamp <- diamonds[sample(nrow(diamonds), 1000), ]
(d <- ggplot(dsamp, aes(carat, price)) +
  geom_point(aes(colour = clarity)) + 
  coord_capped_cart(bottom='none', left='bottom') +
  facet_grid(.~cut) + my.theme
)
```

`facet_grid` and `facet_wrap` have been implemented in versions that display
the axis lines (and labels) on all panels. They work exactly like ggplot2's 
functions, and are named with `_rep_`:
```{r}
d + facet_rep_grid(.~cut)
```

If we want the labels shown as well, use the argument:
```{r}
d + facet_rep_grid(.~cut, repeat.tick.labels = TRUE)
```

It also works for `facet_wrap`:
```{r}
(d2 <- d + facet_rep_wrap(~cut, ncol=2))
```

Finally, the legend can be repositioned to fit in a panel by using 
`reposition_legend`.
See the vignette `legend`. 
The addition of `theme(legend.background)` is merely to provide a border around
the legend.
```{r}
d2 <- d2 + guides(colour=guide_legend(ncol=2)) + 
  theme(legend.background = element_rect(colour='grey'))
reposition_legend(d2, position='center', panel='panel-2-3')
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
