---
title: "Repeat axis lines on facet panels"
author: "Stefan McKinnon Edwards <sme@iysik.com>"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Repeat axis lines on facet panels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,include=FALSE}
library(knitr)
library(ggplot2)

knitr::opts_chunk$set(fig.height=4, fig.width=6,
                      cache=TRUE, autodep = TRUE, cache.path = 'facet-rep-labels/')
```


ggplot2 offers the fantastic option for displaying complex data in forms of
'many multiple', i.e. the facets. From the example of `facet_grid`:

```{r facet_grid_example,fig.cap='Faceting works quite well in its default form. When the panel\'s borders are drawn, nothing lacks.'}
p <- ggplot(mpg, aes(displ, cty)) + geom_point()
p + facet_grid(drv ~ cyl) + theme_bw()
```

In the above example, the panel's borders are drawn with the default settings
of `theme_bw()`.

If we desire the the axis lines, such as those given in this package,
the distinction of the panels disappears.

```{r facet_grid_lemon,fig.cap='The optimised axis lines are gone from inner panels.'}
library(lemon)
p <- p + coord_capped_cart(bottom='both', left='both') +
  theme_bw() + theme(panel.border=element_blank(), axis.line=element_line())
p + facet_grid(drv ~ cyl)
```

The above example is re-created below with both left- and bottom-axis lines repeated.

```{r facet_rep_grid,fig.cap='Axis lines are repeated across all panels by using `facet_rep_grid` of the `lemon` package.'}
library(lemon)
p + facet_rep_grid(drv ~ cyl) + coord_capped_cart(bottom='both', left='both') +
  theme_bw() + theme(panel.border=element_blank(), axis.line=element_line())
```

## Keeping (some) labels

In the following example, we change the facet from a grid to being wrapped on
the interaction of `drv` and `cyl`, *and add free scaling on y-axis*.
`facet_wrap` would normally print the y-axis tick labels for each panel,
but still ignores the x-axis.

```{r facet_wrap,fig.cap="`facet_wrap` keeps y-axis label ticks with `scales='free_y'`."}
p + facet_wrap(~ interaction(cyl, drv), scales='free_y') 
```

A work around by keeping both axes free scale, and fixing the x-axis with either 
`scale_x_continuous` or limits in `cord_*`, but the same x-axis tick labels are
repeated. And this is a bit tedious.

```{r facet_wrap_free,fig.cap='X-axis is entirely fixed, and the plot is littered with x-axis tick labels.'}
p + facet_wrap(~ interaction(cyl, drv), scales='free') + 
  coord_capped_cart(bottom='both', left='both', xlim=c(2,7))
```

We can specify which labels to keep with `facet_rep_wrap`. 
Default is `repeat.tick.labels=FALSE` when `scales='fixed'`
which removes tick labels on all axes (shown in earlier figure).
When using free scales on `facet_rep_wrap`, the appropiate 
labels are drawn.

```{r facet_rep_wrap_left,fig.cap='With `repeat.tick.labels` we are free to specify which sides to keep.'}
p + facet_rep_wrap(~ interaction(cyl, drv), scales='free_y', repeat.tick.labels = 'left')
```

The argument `repeat.tick.labels` accepts a range of arguments: 
`top`,`right`,`bottom`, and `left`, any combination of these are accepted,
as well as
`TRUE` and `FALSE` which are interpreted as all or none,
and `all` and `none`.

## Examples

There are many posibilities. Examples given below (but not executed), and they
might not be pretty.

```{r eval=FALSE}
p + facet_rep_wrap(~ interaction(cyl, drv), scales='free_y', repeat.tick.labels = 'all')
p + facet_rep_wrap(~ interaction(cyl, drv), scales='free_y', repeat.tick.labels = 'bottom')
p + facet_rep_wrap(~ interaction(cyl, drv), scales='free_y', repeat.tick.labels = 'left')
p + scale_x_continuous(sec.axis = dup_axis()) +
  facet_rep_wrap(~ interaction(cyl, drv), scales='free_y', repeat.tick.labels = 'all')
p + scale_x_continuous(sec.axis = dup_axis()) +
  facet_rep_wrap(~ interaction(cyl, drv), scales='free_y', repeat.tick.labels = c('top','left'))
```
