---
title: \pkg{lemon} --- Freshing up your ggplots
author:
  - name: Stefan McKinnon Edwards
    affiliation: The Roslin Insititute, University of Edinburgh
    address:
    - Easter Bush
    - Midlothian
    - EH25 9RG
    - Scotland, UK
    email:  sme@iysik.com
abstract: >
  An abstract of less than 150 words.
preamble: >
  % Any extra latex you need in the preamble
  \usepackage{caption}
  \newcommand{\gtable}{\code{`gtable'}}
  \newcommand{\ggplot}{\code{`ggplot'}}
output: 
  rticles::rjournal_article:
    md_extensions: +latex_macros-tex_math_single_backslash

---


```{r setup,include=FALSE}
library(knitr)
library(ggplot2)
library(lemon)

knitr::opts_knit$set(out.format='latex')
knitr::opts_chunk$set(fig.width=5.5, fig.height=5.5/4*2.5, fig.path='edwards2017-',
                      cache=TRUE, autodep = TRUE, cache.path = 'cache/')
```

\section{Introduction}

The \CRANpkg{ggplot2} package \citep{ggplot2} which utilises the 'grammar of graphics' to
describe a plot is hugely popular. 
Since version 2.0, it has featured an official extension mechanism that has
spawned a number of packages that extends the capabilities of ggplot2.
At the time of writing, \url{ggplot2-exts.org} features 29 packages, 
ranging from themes of popular sources (\CRANpkg{ggthemes}, \citealt{ggthemes}),
to visualisation of graphs (\CRANpkg{ggraphs}, \citealt{ggraph}).

We are adding another package to the list.

Our motivation for this package was to provide *plot polishing* functions.
The package does not supply new geoms or scales, but rather functions 
for improving the look of the final product. 
Examples of these functions are the capped axis lines (\code{cord\_capped\_cart}),
axis brackets (\code{horisontal\_brackets}), or repositioning the guide-box onto 
the main panels, but in exact terms (\code{reposition\_legend}).

Most functions in this package work by manipulating the grobs and gtable objects that
are created when the ggplot plot is to be drawn, so we start this paper by
given a brief discussion of the gtable objects in context of ggplot2.
We then give examples of the funcions introduced in this package.



\section{A brief introduction to gtable}

We assume the reader is familiar with \CRANpkg{ggplot2} \citep{ggplot2}. 
In short, it differs from R's native plotting functions in that we use it to 
*describe* the visualisation we want, rather than instructions of which lines
to draw where.

The majority of the \pkg{lemon} package's functions however works on the 
graphical objects that describe what is being drawn. 
These graphic objects of the \CRANpkg{grid} package, 
or grobs \citep{Murrell2005}, appear after ggplot2 has done 
its magic on translating the grammar of graphics into instructions of what to draw where.

We will only give a brief introduction to grob objects, but refer to 
e.g. \citet{Zhou2010} for a more general introduction, that also includes
the important \dfn{viewports}.
Other authors, such as Baptiste Augui\'{e} who has made considerable contributions
to using ggplot2, has a guide on  at
https://github.com/baptiste/gridextra/wiki/gtable.
We will however focus this introduction on the gtable object returned from ggplot.



The \dfn{gtable} object of the \CRANpkg{gtable} package \citep{gtable} 
has its introduction when a ggplot object is to be drawn on a device. 
Here, the device may be the screen, a plot window, a file, etc.
Calling to view a ggplot object, either explicitly through \code{print} or \code{plot},
or through any other of the multitude of functions, ggplot2 goes through
the motions of training the object's scales on the data to finally produce a 
printable object (see function \code{ggplot\_build} in ggplot2 package).
This object is a grob table -- a gtable object -- that is drawn by the
'grid' package \citep{Murrell2005,Murrell2011} (but you as a user do not necessarily see this).

To get us started, we create a simple scatter plot using ggplot2:

```{r}
library(ggplot2)
p <- ggplot(mtcars, aes(wt, mpg)) + geom_point() + labs(title='A scatter plot')
g <- ggplotGrob(p)
```

The first object, \code{p}, is a ggplot object, and the second object, \code{g}, 
a gtable object. 
Calling to print the objects, implicitly by calling the object from the command 
line of R or directly with \code{print}, we obtain the output given in figure 
\ref{fig:p_object} for the ggplot object, and below / figure \ref{fig:g_object} for the gtable object:

```{r p_object}
print(p)
```
\captionof{figure}{Printing the ggplot object.}
\label{fig:p_object}

```{r g_object}
print(g)
```
  \captionof{figure}{Printing the gtable object.}
  \label{fig:g_object}

The gtable object is of a completely different nature than the 
ggplot object.
Where the ggplot object describes the mapping between variables, geoms, and 
scales, the gtable object positions the plot's features in a table-like structure.
Each feature of the plot, i.e\. the background, an axis, or the actual graph ('panel')
are each displayed on a row.

\subsection{The gtable object}

A gtable object is a hierarchical collection of grob and gtable objects,
placing them in a table-like structure. 
The most important components of each gtable object are 
the \code{layout} data frame and the \code{grobs} list that position and contain the 
plot's features,
in addition to \code{widths} and \code{heights} unit lists that describe the 
column widths and row heights.
These are displayed in figure \ref{fig:gtable_show_names}, superimposed on the plot.

```{r gtable_show_names,echo=FALSE,fig.label='fig:gtable_show_names',fig.cap='Overlaying figure \\ref{fig:p_object} with the layout names and grill, by \\code{gtable\\_show\\_names} and \\code{gtable\\_show\\_grill}.'}
library(lemon)
library(gtable)
library(grid)
s <- gtable_show_grill(p)
grid.draw(gtable_show_names(s, plot=FALSE, rect.gp=gpar(fill='white', col='white', alpha=1/4)))
```

The main feature of interest is \code{panel}, which contains the actual graph.
A seen on figure \ref{fig:g_object}, \code{panel} occurs in the 
`r scales::ordinal(which(g$layout$name == 'panel'))` row, 
and thus `r scales::ordinal(which(g$layout$name == 'panel'))` grob in the grob list.
The three first columns in figure \ref{fig:g_object} are compiled from 
the \code{layout} data frame and is for the panel,


```{r}
subset(g$layout, name == 'panel')
```

A grob can be drawn across multiple cells, covering both multiple rows and / or
multiple rows.
This span is specified by the top-left cell and bottom-right cell, 
corrosponding to the \code{t}, \code{l}, \code{b}, and \code{r} coordinates above.
Grobs can also drawn on top of each other, starting with lowest z-index, as given 
by the \code{z} column.
When 'clipping' is enabled, by setting \code{clip} to \code{'on'}, 
anything drawn outside the spanned cells is hidden.
The \code{name} column allows us to locate the grobs for futher manipulation.


We mention shortly here, that a grob in this context can be a simple grob primitive 
such as a rectangle (see the background) or a \code{'zeroGrob'} (a blank placeholder),
or a collection of grobs such as the panel and axes (\code{'absoluteGrob'} and \code{'titleGrob'}
are both specialised classes of grid's \code{'gTree'}). 

For \code{panel}, the grob too is a collection of grobs, in the form of a \code{gTree}
with a list of children grobs:

```{r}
str(g$grobs[[which(g$layout$name == 'panel')]], max.level=1)
```


Although the order of the layout data frame and grobs list are 
intrinsically linked, they are not enforced. 
The layout data frame can be accessed via \code{g\$layout} and the grobs list via
\code{g\$grobs}, and they can be manipulated separately as you 
normally would data frames and lists, respectively. 


\subsection{Widths and heights}
The gtable is also described by the heights and widths of rows and columns. 
These two components are each a 'unit.list' with a component for each row or 
column.
A quick glance at the widths of \code{g} tells us a lot:

```{r}
g$widths
```

These can be compared to the columns of figure \ref{fig:gtable_show_names};
the first and last are the plot's margins, given with \code{theme(plot.margin)}.
The second width, \code{1grobwidth}, refers to the actual width of the y-axis title
'mpg'. The third width, \code{sum(1grobwidth, 2.75pt)}, is literally the width \emph{plus}
2.75 pt, with the width here being those of the y-axis labels. 
We will in general see grob-widths and grob-heights for text elements.
The fourth width, \code{1null}, simply expands into the available space and is here 
used to maximise the width of the actual graphing area.
The two remaining widths of 0 cm are those for a right-hand side axis, which can 
be seen on figure \ref{fig:gtable_show_names}, where their respective columns have been
expanded for us to view.

The elements of widths, and similarily for heights, are not fixed to those shown here.
Adding a legend or using facetting will change them.


\subsection{Manipulating the gtable object}

The \CRANpkg{gtable} package supplies a number of functions for manipulating
a gtable object, such as adding rows and columns (\code{gtable\_add\_rows},
\code{gtable\_add\_cols}), combining several gtable objects together as if 
they were matrices (\code{cbind}, \code{rbind}),
or adding new features to the gtable object (\code{gtable\_add\_grob}).

These functions are usually sufficient for manipulating the gtable object.
Knowing the components of the gtable object however allows us to manipulate 
the gtable object in an alternative and direct fashion.

\subsubsection{Adding or replacing features}

For adding a grob, \pkg{gtable} comes with the function 
\code{gtable\_add\_grob(x, grobs, t, l, b, r, z, clip, name)}.
This adds a new row to the \code{x\$layout} data frame with the coordinates of the
spanning cells, a \code{name} for the grob, 
and appends \code{grobs} to the \code{x\$grobs} list.

\emph{Replacing a grob} is however more tricky. 
To do so, we have to work out which index in the \code{grobs} list corresponds 
to the feature that is to be replaced. 
The following example demonstrates this and then replaces the x-axis label with 
a pink box.

```{r}
g <- ggplotGrob(p)
i <- which(g$layout$name == 'xlab-b')
g$grobs[[i]] <- roundrectGrob(r=unit(0.5, 'snpc'), gp=gpar(fill='pink', col='black'))
grid.draw(g)
```
\captionof{figure}{Replacing the x-axis label grob with a pink square.}
\label{fig:replace_grob}


One could naturally just have added the grob with \code{gtable\_add\_grob},
placing it on top of the x-axis label. But if the grob was not filled with a 
solid colour, the label would still appear.
The example however also illustrates how to \emph{extract} a grob, if the grob
needed altering.

\emph{Row heights} and \emph{column widths} can be directly altered by changing the element
of \code{g\$heights} or \code{g\$widths}. 
Although these are specified to fit the plot's features while maximising the 
size of the panels, we can change them.
To change the width of the panel we must change the width of the column it resides in.

In the following example, we find the row in the layout data frame corresponding to the panel.
The relevant column is then specified by the \code{l} coordinate,
and we can proceed to update the widths.

```{r}
g <- ggplotGrob(p)
layout <- subset(g$layout, name == 'panel')
g$widths[[layout$l]] <- unit(5, 'cm')
grid::grid.draw(g)
```

As \code{g\$widths} is a \code{unit.list} object, it requires using double square brackets
for replacing an element. This also implies changing the widths must be done
one by one, or by building a new \code{`unit.list'} object entirely.

\section{Demonstration}

The \pkg{lemon} package has functions for working with the legend (or 'guide-box') and axis lines.

\subsection{Legends}

The primary contribution here is the function \code{reposition\_legend}.
The default behaviour for \CRANpkg{ggplot2} is to place the legend in one of
the four margins of the plot.
Repositioning the legend to occur within the panel can be done by supplying 
a two-element numeric vector, specifying the position \emph{relative to the 
entire plot}, to \code{theme(legend.position)}.
This has the unfortunate effect that the legend is not positioned correctly after
changing any size of the plot (e.g.\ resizing the plot's size or changing the font sizes).

The function \code{reposition\_legend} allows us to position the relative to
any cell of the gtable. In order to do so, it forces the plot into a \gtable
object, which is always returned.

```{r reposition_legend}
reposition_legend(p %+% aes(color=cyl), position = 'top right')
```
\captionof{figure}{Repositioning the legend on to the main panel.}
\label{fig:reposition_legend}

The appearences of the legend is not modified in any way, thus modifying it is
done by the usual methods of \CRANpkg{ggplot2}. These could be
\code{theme(legend.direction)} or \code{guide\_legend(...)}.
We also highlight \code{theme(legend.box.background)} when having 
multiple legends as the component \code{theme(legend.background)} only affects
the background of each singular legend, not the area spanning all legends.

The function can also be used to add multiple legends to different locations,
as demonstrated in figure \ref{fig:multiple_legends}.

```{r multiple_legends}
x <- p %+% aes(color=cyl, size=qsec)
g <- g_legend(x)
x <- reposition_legend(x, position='bottom left', legend=g$grobs[[2]], plot=FALSE)
x <- reposition_legend(x, position='top right', legend=g$grobs[[1]], plot=FALSE)
grid.draw(x)
```
\captionof{figure}{Repositioning multiple legends to different locations.}
\label{fig:multiple_legends}

The above code also demonstrates the usage of \code{g\_legend} to extract the 
legend from a plot.
We must however emphasise that \code{g\_legend} was first proposed by
Baptiste Augui\'{e} as early as June 2012 on \CRANpkg{ggplot2}'s 
Github wiki\footnote{\url{https://github.com/tidyverse/ggplot2/wiki/Share-a-legend-between-two-ggplot2-graphs/047381b48b0f0ef51a174286a595817f01a0dfad}}.
Baptiste Augui\'{e} is the author of \CRANpkg{gridExtra} \citep{gridExtra}.
It has since propogated throughout a swarm of 
Stack Overflow answers.
For this same reason, Baptiste Augui\'{e} has been added as contributor to this
package.

Lastly, a modified version of \code{grid.arrange} with a shared legend is 
demonstrated in figure \ref{fig:brackets}.
The function \code{grid\_arrange\_shared\_legend} supplied in this package
is modified from Shaun Jackman's 
version\footnote{\url{http://rpubs.com/sjackman/grid\_arrange\_shared\_legend}} 
and Baptiste Augui\'{e} 
version\footnote{\url{https://github.com/tidyverse/ggplot2/wiki/Share-a-legend-between-two-ggplot2-graphs}}.

\subsection{Axis lines and brackets}


The second domain of functions in \pkg{lemon} is for axis lines and brackets.
We desired to alter the axis lines to be capped to tick marks, or even replaced
by brackets.

In the following example, three functions are demonstrated: 
brackets, capped axis lines, 

```{r brackets}
p <- ggplot(mpg, aes(as.factor(cyl), hwy, colour=class)) +
  geom_point(position=position_jitter(width=0.3)) +
  theme_bw()

# Demonstrate brackets
p1 <- p + coord_flex_cart(bottom=brackets_horisontal(length=unit(0.08, 'npc'))) +
  theme(panel.border = element_blank(), axis.line = element_line(),
        panel.grid.major.x = element_blank())

p <- ggplot(mpg, aes(displ, hwy, colour=class)) + geom_point() +
  theme_bw()

# Demonstrate capped axis lines
p2 <- p + coord_capped_cart(bottom='right', left='none', gap = 0.03) +
  theme(panel.border = element_blank(), axis.line = element_line(),
        panel.grid.major.x = element_blank())


# Demonstrate shared legend
grid_arrange_shared_legend(p1, p2, position='bottom')
```
\captionof{figure}{
  \emph{Left:} Displaying brackets on an axis to indicate categorical values.
  \emph{Right:} Capped axis lines, with x-axis capped at right-most tick, and
               y-axis capped by a small proportion.
}
\label{fig:brackets}

\emph{To do:} Explain features of brackets (direction, height, style taken from theme),
and capped lines (coords, flexible coords), added theme for polishing, etc.


The use of axis lines to be more informative of the displayed data has been
shown in \citep[p.\ 130]{Tufte2001} with \dfn{range-frames}, where axis lines
are limited to the range of the displayed data.
This has been implemented in \CRANpkg{ggthemes} \citep{ggthemes} with
\code{geom_rangeframe}.

Our capped axis lines differs from these by intending to display the valid range
the values may take. E.g.\ correlations where the value cannot exceed \pm 1.

\CRANpkg{ggplot2}'s facets however avoid displaying redundant axis lines and 
labels, when the scale is identical along a row or column.



\section{Summary}


\subsection{Acknowledgements}

This article was initiated with the `rticles` package \citep{rticles}.

The author, SME, was funded by the MRC (FARSPhase: a Flexible, widely Applicable, Robust, and Scalable phasing algorithm for human genetics, grant no. MR/M000370/1),
while working on this software.

\href{https://baptiste.github.io/}{Baptiste Augui\'{e}} for his work on many
functions for \CRANpkg{ggplot2} and \CRANpkg{grid} for which without, 
this package could not have been possible.


This file is only a basic article template. For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf).

\bibliography{edwards2017}
