---
title: Lemon --- yet another ggplot2 extension package
author:
  - name: Stefan McKinnon Edwards
    affiliation: The Roslin Insititute, University of Edinburgh
    address:
    - Easter Bush
    - Midlothian
    - EH25 9RG
    - Scotland, UK
    email:  sme@iysik.com
abstract: >
  An abstract of less than 150 words.
preamble: >
  % Any extra latex you need in the preamble
output: rticles::rjournal_article
---


```{r setup,include=FALSE}
library(knitr)
library(ggplot2)

knitr::opts_chunk$set(fig.height=4, fig.width=6,
                      cache=TRUE, autodep = TRUE, cache.path = 'cache/')
```

## Introduction

Motivation



### A brief introduction to gtable

We assume the reader is familiar with ggplot2 \citep{ggplot2}.
In short, it differs from R's native plotting functions in that it only 
*describes* the visualisation.
The ggplot object returned from calling `ggplot(data, aes(...)) + ...`
only describes the mapping between variables, geoms (points, bars, etc.),
and scales, and the theme.

The 'gtable' package \citep{gtable} and object has its introduction when a ggplot 
object is to be drawn on a device. 
Here, the device may be the screen, a plot window, a file, etc.
Calling to view a ggplot object, either explicitly through `print` or `plot`, 
or through any other of the multitude of functions, ggplot2 goes through
the motions of training the object's scales on the data to finally produce a 
printable object (see function `ggplot_build` in ggplot2 package).
This object is a grob table -- a gtable object.

The gtable object is of a completely different nature than the 
ggplot object.
Where the ggplot object describes the mapping between variables, geoms, and 
scales, the gtable object describes which lines are drawn where. 
It does so by using graphical objects -- 'grobs' -- 
as introduced in the package 'grid' by Paul Murrell.
We will only give a brief introduction to grob objects, but refer to 
e.g. \citet{Zhou2010} for a more general introduction, that also includes
the important 'viewports'.
Other authors, such as Baptiste Auguié who has made considerable contributions
to using ggplot2, has a guide at
https://github.com/baptiste/gridextra/wiki/gtable.


A gtable object is a hierarchical collection of grob and gtable objects,
placing them in a table-like structure. 
The two important components of each gtable object are 
the `layout` data frame and the `grobs` list.

Before continuing our exploration of the gtable object, we start with an example
that demonstrates the different between the ggplot object and the gtable object.

```{r}
p <- ggplot(mtcars, aes(wt, mpg)) + geom_point()
```

```{r overlay}
overlay <- gpar(colour='black')

g <- ggplotGrob(p)
grid.draw(g)
y <- unit(0, 'npc')
for (i in 0:(nrow(g))) {
  grid.lines(x=c(0,1), y=y, gp=overlay)
  y <- y + g$heights[i]
}
ys <- 
hlines <- polylineGrob(x=rep(c(0,1), each=nrow(g)+1), y=unit.c(unit(0,'npc'), g$heights, unit(0,'npc'), g$heights), id = rep(1:nrow(g)+1, times=2), gp = overlay )
grid.draw(hlines)

```

To understand a `gtable` object, 






## Section title in sentence case

This section may contain a figure such as Figure \ref{figure:rlogo}.

\begin{figure}[htbp]
  \centering
  \includegraphics{Rlogo}
  \caption{The logo of R.}
  \label{figure:rlogo}
\end{figure}

## Another section

There will likely be several sections, perhaps including code snippets, such as:

```{r}
x <- 1:10
x
```

## Summary

## Acknowledgements

This article was initiated with the `rticles` package \citep{rticles}.


This file is only a basic article template. For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf).
\bibliography{RJreferences}
