---
title: Lemon --- yet another ggplot2 extension package
author:
  - name: Stefan McKinnon Edwards
    affiliation: The Roslin Insititute, University of Edinburgh
    address:
    - Easter Bush
    - Midlothian
    - EH25 9RG
    - Scotland, UK
    email:  sme@iysik.com
abstract: >
  An abstract of less than 150 words.
preamble: >
  % Any extra latex you need in the preamble
output: rticles::rjournal_article
---


```{r setup,include=FALSE}
library(knitr)
library(ggplot2)

knitr::opts_chunk$set(fig.height=4, fig.width=6,
                      cache=TRUE, autodep = TRUE, cache.path = 'cache/')
```

## Introduction

Motivation



### A brief introduction to gtable

We assume the reader is familiar with ggplot2 \citep{ggplot2}. 
In short, it differs from R's native plotting functions in that it only
*describes* the visualisation. 
The ggplot object returned from calling 
`ggplot(data, aes(...)) + ...` only describes the mapping between variables, 
geoms (points, bars, etc.), and scales, and the theme.


The 'gtable' package \citep{gtable} and object has its introduction when a ggplot 
object is to be drawn on a device. 
Here, the device may be the screen, a plot window, a file, etc.
Calling to view a ggplot object, either explicitly through `print` or `plot`, 
or through any other of the multitude of functions, ggplot2 goes through
the motions of training the object's scales on the data to finally produce a 
printable object (see function `ggplot_build` in ggplot2 package).
This object is a grob table -- a gtable object.

The gtable object is of a completely different nature than the 
ggplot object.
Where the ggplot object describes the mapping between variables, geoms, and 
scales, the gtable object describes which lines are drawn where. 
It does so by using graphical objects -- 'grobs' -- 
as introduced in the package 'grid' by Paul Murrell.
We will only give a brief introduction to grob objects, but refer to 
e.g. \citet{Zhou2010} for a more general introduction, that also includes
the important 'viewports'.
Other authors, such as Baptiste Augui\'{e} who has made considerable contributions
to using ggplot2, has a guide at
https://github.com/baptiste/gridextra/wiki/gtable.
We will however focus this introduction on the gtavle object returned from ggplot.


A gtable object is a hierarchical collection of grob and gtable objects,
placing them in a table-like structure. 
The two important components of each gtable object are 
the `layout` data frame and the `grobs` list.

Before continuing our exploration of the gtable object, we start with an example
that demonstrates the different between the ggplot object and the gtable object.
Below is a piece of code that creates a plot.
`p` is a ggplot object and when called in an interactive R session, 
`print.ggplot(p)` is called to display the graphics. 
This function creates the gtable object which is drawn by passing it to 
`grid.draw` of the 'grid' package. 
The same functionality can be obtained by calling `ggplotGrob`.
The difference between the ggplot and gtable object is clear when we print the 
latter:

```{r}
p <- ggplot(mtcars, aes(wt, mpg)) + geom_point()
g <- ggplotGrob(p)
g
```


The table structure is revealed, displaying the two important components, the 
layout and the grob list. 
Each grob (or a collection of grobs) appears on each row in conjunction with 
which cells they span. For example, the first grob (`background`) spans the 
first to tenth row, counting from the top, 
and first column to seventh column, counting from left. 
The grob that contains the actual graph (`panel`) occupies just the cell in the 
sixth row and fourth column. 
The panel is however drawn *on top* of the background, as
indicated by the z-index. 


The main component of interest is `panel`, which contains the actual plot. 
The background component is a simple rectGrob (`rect[..]`), 
The grob descriptor (`gTree[..]`) tells us that this is a collection of grobs, 
as is the left-hand axis ('`axis-l`', `absoluteGrob[..]`). 
They are both grob lists, but the latter differs from the former by having
a height and width defined.
The axis line, ticks, and labels, a child grob of the left-hand axis grob, is
itself a gtable. 
Finally, many components are of the type `zeroGrob`, 
which are essentially the grobs' version of `NULL`, a blank placeholder. 



The order of the layout data frame and grobs list are intrinsically linked, but not enforced. The layout data frame can be accessed via `g$layout` and the grobs list via `g$grobs`, and they can be manipulated separately. The The grob descriptors, shown in the last column of output 1 are thus picked up only when printing this view.

The gtable is also described by the heights and widths of rows and columns. 


We utilise the functions `gtable_show_grill` and `gtable_show_names` in 
succession to visualise the elements.

```{r}
library(grid)
grill <- gtable_show_grill(p)
grid.draw(gtable_show_names(grill, plot=FALSE))
```


## Section title in sentence case

This section may contain a figure such as Figure \ref{figure:rlogo}.

\begin{figure}[htbp]
  \centering
  \includegraphics{Rlogo}
  \caption{The logo of R.}
  \label{figure:rlogo}
\end{figure}

## Another section

There will likely be several sections, perhaps including code snippets, such as:

```{r}
x <- 1:10
x
```

## Summary

## Acknowledgements

This article was initiated with the `rticles` package \citep{rticles}.


This file is only a basic article template. For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf).
\bibliography{RJreferences}
